Cognate beginnings to bilingual lexical acquisition

### 2023-06-13 manuscript/manuscript.qmd
Purge dependencies

format:
  pdf:
    papersize: letter
    fontsize: 12pt
    linestretch: 1.5
    date-format: full
    template-partials:
    - templates/title.tex
    - templates/before-body.tex
    include-in-header:
    - templates/include-in-header.tex
    header-includes: 
    - \usepackage{tipa}
    - \usepackage{nowidow}
    geometry:
    - top=1in
    - bottom=1in
    - left=1in
    - right=1in
    colorlinks: true
    linkcolor: DodgerBlue
    urlcolor: DodgerBlue
    citecolor: DodgerBlue
    link-citations: true
    biblio-config: false
    cite-method: biblatex
    biblatex-chicago: false
    biblio-style: apa
    biblatexoptions: backend=biber
    block-headings: false
    indent: true

author-note:
  disclosures:
    conflict-of-interest: |
        The authors declare no conflicts of interest with regard to the funding source of this study.
    financial-support: |
        This research was supported by grants from the Spanish Ministerio de Ciencia, Innovación y Universidades (PGC2018-101831-B-I00 and PRE2019-088165), and the Catalan Government [ICREA (Catalan Institution for Research and Advanced Studies) Academia 2019 award]. Gonzalo Garcia-Castro was supported by a fellowship of the Spanish Ministerio de Ciencia, Innovación y Universidades (FPI 2019).
    gratitude: |
        We are grateful to Chiara Santolin, Ege E. Özer, Alicia Franco-Martínez, Cristina Rodríguez-Prada, and the rest of the Speech Acquisition and Perception research group for their helpful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support, and Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We also thank all families and infants who participated in the experiments.

abstract: |
    Language non-selectivity is a prominent feature of bilingual lexical processing. An instance of such non-selectivity is embodied by cognateness. Recent studies have pointed to cognateness as a facilitator of vocabulary growth, and to cognates being acquired at earlier ages than non-cognates. The specific mechanisms behind such facilitation are unclear. Building on accumulator models of language acquisition, we suggest that cognateness increases the number of cumulative learning instances that a bilingual child encounters, thanks to cross-language activation. In line with previous studies, this model predicted a cognate facilitation, but only for words to which children are exposed less often. We tested these predictions against vocabulary data from 366 Catalan-Spanish bilinguals toddlers. We used Bayesian Explanatory Item Response Models to estimate participants’ probability of acquisition of 604 words, conditional to the cognate status and lexical frequency of the word-form, and the age and degree of exposure to each language of the participant. We found a cognateness facilitation effect only in low-frequency words or words from the language of lower exposure. Overall, our findings support an interactive account of bilingual vocabulary acquisition in which the lexical representations in one language interact with the acquisition of words in the other language.

@@

Consider the example of the Catalan-Spanish cognate translation equivalent /\textipa{"gat}-\textipa{"ga.to}/ [cat], whose word-forms share three phonemes (0.75 Levenshtein similarity). When the child is exposed to /\textipa{"gat}/, they might co-activate /\textipa{"ga.to}/ in parallel, due to their phonological similarity. If this exposure to the word form occurs in a convenient referential context (e.g., in the presence of a cat), this might count as a learning instance for both active word-forms. Consider now the case of the non-cognate translation equivalent /\textipa{"gos}-\textipa{"pe.ro}/. Because of their low phonological similarity (0% Levenshtein similarity), the amount of co-activate of both words might be low, which might leads to such exposure counting as a learning instance for /\textipa{"gos}/ (which the child was exposed to), but not for  /\textipa{"pe.ro}/. While /\textipa{"gat}-\textipa{"ga.to}/ are predicted to benefit from parallel activation, /\textipa{"gos}-\textipa{"pe.ro}/ are not. This facilitation effect may not be equivalent for /\textipa{"gat}-\textipa{"ga.to}/. If the child receives a larger amount of Catalan input than Spanish input, they will encounter the Catalan form /\textipa{"gat}/ more frequently than the Spanish form /\textipa{"ga.to}/. If parallel activation is in place, the former will activate the latter more often than vice versa. Ultimately, /\textipa{"ga.to}/ should benefit more strongly from its cognate status than /\textipa{"gat}/. For instance, if the previous child is exposed 80% of the time to Catalan, and 20% of the time to Spanish, words in Spanish should, on average, show an even stronger cognateness facilitation effect than words in Catalan.

@@

The present study contributes a specific data point into the complex landscape of bilingualism research. Bilinguals are a remarkably heterogeneous population that can hardly be satisfactorily characterised in a comprehensive way [@sebastian-galles2020bilingual]. Bilinguals differ across multiple dimensions. these differences span from exclusively linguistic factors such as the amount of overlap between the phonemic inventories of the two languages being learned (e.g., low, like the case of English and Mandarin, or high like the case of Spanish and Greek), to higher-level factors like the socio-linguistic situation in which the two languages co-exist (e.g., in some regions both languages are co-official and used in similar contexts, while in others, one of the languages hardly has any societal presence, i.e., heritage languages). This diversity of situations in which bilingual toddlers acquire language calls for special consideration of the generalisability of results in bilingualism research. Our sample, although homogeneous (e.g., similar parental educational level across), represents a fairly particular population of bilinguals: the languages involved in the present investigation, Catalan and Spanish, co-exist in Catalonia as official languages, both languages are used in fairly similar contexts, and both languages are known by the majority of the population. In 2018, more than 81.2% of a representative sample of 8,780 adults aged 15 years or older living in Catalonia reported being able to speak Catalan, and more than 99.5% of the same population reported being able to speak Spanish [@2018els]. In addition, and as commented in the introduction, Catalan and Spanish are both Romance languages and share a considerable amount of cognates. Extending our analyses to other bilingual populations learning typologically more distant languages, and whose languages tend to be used in more distinct contexts (e.g., heritage languages) should be a natural future step for the present investigation.

>>>>

format:
  pdf:
    papersize: letter
    fontsize: 12pt
    linestretch: 1.5
    date-format: full
    template-partials:
    - templates/title.tex
    - templates/before-body.tex
    include-in-header:
    - templates/include-in-header.tex
    header-includes: 
    - \usepackage{tipa}
    - \usepackage{nowidow}
    geometry:
    - top=1in
    - bottom=1in
    - left=1in
    - right=1in
    colorlinks: true
    linkcolor: DodgerBlue
    urlcolor: DodgerBlue
    citecolor: DodgerBlue
    link-citations: true
    biblio-config: false
    cite-method: biblatex
    biblatex-chicago: false
    biblio-style: apa
    biblatexoptions: backend=biber
    block-headings: false
    indent: true
  docx:
    knitr:
      opts_chunk:
        dev: "png"
        dev-args:
          - "cairo-png"
        dpi: 300
        out.width: "80%"
author-note:
  disclosures:
    conflict-of-interest: |
        The authors declare no conflicts of interest with regard to the funding source of this study.
    financial-support: |
        This research was supported by grants from the Spanish Ministerio de Ciencia, Innovación y Universidades (PGC2018-101831-B-I00 and PRE2019-088165), and the Catalan Government [ICREA (Catalan Institution for Research and Advanced Studies) Academia 2019 award]. Gonzalo Garcia-Castro was supported by a fellowship of the Spanish Ministerio de Ciencia, Innovación y Universidades (FPI 2019).
    gratitude: |
        We are grateful to Chiara Santolin, Ege E. Özer, Alicia Franco-Martínez, Cristina Rodríguez-Prada, and the rest of the Speech Acquisition and Perception research group for their helpful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support, and Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We also thank all families and infants who participated in the experiments.

abstract: |
    Recent work on bilingual language acquisition points to cognate words being acquired earlier in age than non-cognates. Cross-language activation has been proposed as a candidate mechanism behind this facilitation effect, but the

       but the specific dynamics of lexical Building on accumulator models of language acquisition, we present an account of word acquisition in which the co-activation of cognate translation equivalents increases the rate at which of bilingual children encounter learning instances for words in both languages. We tested these predictions against vocabulary data from 366 Catalan-Spanish bilinguals toddlers. We used Bayesian Explanatory Item Response Models to estimate participants’ probability of acquisition of 604 words, conditional to the cognate status and lexical frequency of the word-form, and the age and degree of exposure to each language of the participant. We found a cognateness facilitation effect only in low-frequency words or words from the language of lower exposure. Overall, our findings support an interactive account of bilingual vocabulary acquisition in which the lexical representations in one language interact with the acquisition of words in the other language.

@@

Consider the example of the Catalan-Spanish cognate translation equivalent /\textipa{"gat}-\textipa{"ga.to}/ [cat], whose word-forms share three phonemes (0.75 Levenshtein similarity). When the child is exposed to /\textipa{"gat}/, they might co-activate /\textipa{"ga.to}/ in parallel, due to their phonological similarity. If this exposure to the word form occurs in a convenient referential context (e.g., in the presence of a cat), this might count as a learning instance for both active word-forms. Consider now the case of the non-cognate translation equivalent /\textipa{"gos}-\textipa{"pe.ro}/. Because of their low phonological similarity (0% Levenshtein similarity), the amount of co-activation of both words might be low, which might leads to such exposure counting as a learning instance for /\textipa{"gos}/ (which the child was exposed to), but not for  /\textipa{"pe.ro}/. While /\textipa{"gat}-\textipa{"ga.to}/ are predicted to benefit from parallel activation, /\textipa{"gos}-\textipa{"pe.ro}/ are not. If the child receives a larger amount of Catalan input than Spanish input, they will encounter the Catalan form /\textipa{"gat}/ more frequently than the Spanish form /\textipa{"ga.to}/. If parallel activation is in place, the former will activate the latter more often than vice versa. Ultimately, /\textipa{"ga.to}/ should benefit more strongly from its cognate status than /\textipa{"gat}/. For instance, if the previous child is exposed 80% of the time to Catalan, and 20% of the time to Spanish, words in Spanish should, on average, show an even stronger cognateness facilitation effect than words in Catalan.

@@

The present study contributes a specific data point into the complex landscape of bilingualism research. Bilinguals are a remarkably heterogeneous population that can hardly be satisfactorily characterised in a comprehensive way [@sebastian-galles2020bilingual]. Bilinguals differ across multiple dimensions. These differences span from exclusively linguistic factors such as the amount of overlap between the phonemic inventories of the two languages being learned (e.g., low, like the case of English and Mandarin, or high like the case of Spanish and Greek), to higher-level factors like the socio-linguistic situation in which the two languages co-exist (e.g., in some regions both languages are co-official and used in similar contexts, while in others, one of the languages hardly has any societal presence, i.e., heritage languages). This diversity of situations in which bilingual toddlers acquire language calls for special consideration of the generalisability of results in bilingualism research. Our sample, although homogeneous (e.g., similar parental educational level across), represents a fairly particular population of bilinguals: the languages involved in the present investigation, Catalan and Spanish, co-exist in Catalonia as official languages, both languages are used in fairly similar contexts, and both languages are known by the majority of the population. In 2018, more than 81.2% of a representative sample of 8,780 adults aged 15 years or older living in Catalonia reported being able to speak Catalan, and more than 99.5% of the same population reported being able to speak Spanish [@2018els]. In addition, and as commented in the introduction, Catalan and Spanish are both Romance languages and share a considerable amount of cognates. Extending our analyses to other bilingual populations learning typologically more distant languages, and whose languages tend to be used in more distinct contexts (e.g., heritage languages) should be a natural future step for the present investigation.
